Seismic velocity estimation and the Canadian crust

### 2014-01-28 paper/bpostlethwaite_Thesis.tex
thesis submission

\documentclass[review]{elsarticle}
\usepackage{graphicx,multirow,color,natbib,amssymb}
\journal{Tectonophysics}

@@

\title{Seismic parameter estimation and the Canadian crust}

@@

\address[ubc]{Earth and Ocean Sciences, UBC, Vancouver BC, Canada}
\address[gsc]{Geological Survey Canada, Natural Resources Canada, Ottawa ON, Canada}
%% ------------------------------------------------------------------------ %%
\begin{abstract}
It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurments are compared with results for matching stations from previous studies. Analysis of results in conjunction with additonal velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.
\end{abstract}

@@

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and sourthern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occured as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale décollement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and has enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of Eaton (2005) who considers seismic observations in light of several hypotheses for the origin of the continental Moho: relic Moho, underplating, metamorphic front and regional d\'ecollement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis whereas the presence of high-velocity, lower-crustal xenoliths in southern Alberta favours thick mafic underplating. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lends support to the third hypothesis, that at least in some places, continental Moho forms by large-scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections separated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional d\'ecollement. A subsequent study by Cook et al. (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration.

In another recent summary study of Lithoprobe data, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust and note that Archean cratons and their boundaries contain structures similar to those found in Proterozoic and Phanerozoic orogens. They also discuss the lack of evidence for significant crustal disruption from magmatic instrusion related to subduction, post-orogenic extension or orogenic collapse. The authors remark upon the striking uniformity in Moho depth that is imaged between 33km and 43km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both Cook et al. (2009) and Hammer and Clowes (2011) raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary.

The Lith5.0 revised crustal model presented by Perry, Eaton and Forte (2002) was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick ($>$42.5km) Phanerozoic crust in the south and west of the continental craton.

In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 2000), and that the latter quantity provides some ability to discriminate among possible bulk compositions (Zandt and Ammon, 1995). The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described in (Bostock et al, 2010). Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen (Eaton et al., 2005), a structural and compositional characterization of the Superior Province (Darbyshire et al. 2006) and a crustal evolution analysis of the Hudson Bay region (Thompson et al., 2010). Results from these studies will serve as benchmarks for our expanded effort. In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian Crust.

@@

Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source is teleseismic receiver function (hereafter RF) analysis. The remaining two datasets are Crust 1.0, a widely published statistical average of similar regions with a global two degree resolution, and a compilation of pre-processed active source data (Mooney, 2012). Both Crust 1.0 and the active source compilation are used to qualify and compare with the primary dataset.

\subsection{Teleseismic Data Set}
The primary data utilized in this study are computed from teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30 to 100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 events are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

@@

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{reflectedPhases.pdf}
  \caption{Schematic diagram illustrating geometry of phases for the velocity contrast representing the Moho}
  \label{fig:reflectedPhases}
\end{figure}

A well tested and widely published method for extracting the ratio of {\it P} velocity $V_P$ to {\it S} velocity $V_S$, herafter referred to as $R$, and crustal thickness (or depth to Moho) $H$, is outlined by Zhu and Kanamori (2000), hereafter ZK. This method exploits the dependence of these parameters to the differential arrival times between the {\it S} wave reflected phases $Ps$, $PpPs$, $PsPs$, $PpSs$ and $PsSs$ and the direct {\it P} wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ are kinematic analogues, such that the energy for these two phases arrive simultaneously for a 1-D model. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

@@

The travel time equations, as employed by ZK, assume a value for crustal {\it P} wave velocity, $V_P$, that in practice will trade-off to some degree with crustal thickness $H$. In our implementation of the ZK approach, each station is assigned a $V_P$ value corresponding to the Crust 1.0 value for the corrisponding $2^\circ$ containing cell. In the specific cases where data are being compared to previously published results for quality control, $V_P$  values are chosen to match the values in the published study.

@@

A limitation of the ZK approach is the requirement of an initial $V_P$ estimate. In principal we may compute the stacking function Eq. \ref{eq:stack}, as a function of all three seismic parameters, $s(H,R,V_P)$. Taking advantage of modern multi-core systems with reasonable quantities of data (fifty to a few hundred RFs) and a moderate search space ($n^3$ parameter candidates where $n \approx 150$) the computational time makes this method easily scalable to hundreds of stations. This ``fullgrid'' approach is analyzed to determine the viability of recovering $V_P$.

@@

The Crust 1.0 model (Laske et. al., 2013) employs many of these same active source data to provide regular, complete coverage across all of Canada. The model includes estimates of $V_P$ and $V_S$ information with depth at $1^\circ$ resolution in latitude and longitude. In addition to employing local active source data, Crust 1.0 is also constrained by statistically aggregated data from geographically distant but geologically similar regions. It therefore affords a somewhat independent dataset suited for comparison with the crustal estimates calculated in this study.

@@

Three published studies have provided parameter estimates for $R$ and $H$ for subsets of the seismic stations considered here. The most recent study (Thompson et. al., 2010) focused on the northern Canadian Shield and employs both the linear stacking approach used by ZK as well as a phase weighted method. A 2007 study of the Superior Province (Darbyshire et. al, 2007) and a study of the Grenville Orogen (Eaton et. al., 2006) utilize the ZK + semblance approach outlined in section \ref{section:VpVsMethod}.  Comparisons between crustal estimates of $R$ for particular seismic stations are made for those values with corresponding standard errors of less than $\pm$0.06. This error threshold is chosen empirically as that which best partitions those receiver function stacks with visually recognizable energy along phase moveout curves from those where lack of data or poor quality make these moveout curves difficult to distinguish. For each study under comparison, data are reprocessed using the $V_P$ chosen by the study authors. For the 2006 and 2007 studies the ZK + semblance method is used whereas for the Thompson et. al study the linear stacking ZK approach is employed. Both correlation coefficient and mean difference are provided in (Table \ref{table:comparison}) as measures of similarity.

\begin{table}
  \begin{tabular}{ p{3cm} p{2cm} p{1cm} p{0.5cm} p{0.5cm} p{0.5cm} p{0.5cm}}
    & & & \multicolumn{2}{ c }{Corr. Coeff.} & \multicolumn{2}{ c }{RMS difference} \\
    \hline
    Study Authors & Region & Num. Stns &  $H$ & $R$  &$H$ (km)& $R$ \\
    \hline
    Thompson et. al. (2010) & N. Canadian Shield & 29 & 0.97 & 0.70 & 0.78 & 0.017 \\
    Darbyshire et. al. (2007) & Superior Province & 10 & 0.95 & 0.43 & 1.71 & 0.037 \\
    Eaton et. al. (2006) & Grenville Orogen & 26 & 0.90 & 0.61 & 1.03 & 0.033 \\
    \hline
  \end{tabular}
  \caption{Comparison of $R$ and $H$ estimates with three published studies}
\label{table:comparison}

@@

All three studies reported $H$ estimates that correlate strongly with the crustal thickness parameters computed here. The data from the northern Canadian Shield shows strong correlation in both $H$ and $R$. The lower correlation observed in estimates of $R$ than $H$ is not unexpected as the range of $R$ is relatively narrow and small changes ($\approx 2\%$) are physically meaningful. A corollary of this is greater sensitivity of $R$ to differences in procedure such as the choice of deconvolution algorithm and selection and scaling of data. The results from Thompson et. al. (2010) also corrispond most closely to the parameter estimates computed here, with the mean difference for both parameters being less than half of the averaged bootstrap error computed in this study. Results published by {\it Eaton et. al.} (2006) show reasonable correlations in both $R$ and $H$ and a mean difference for both parameters that is roughly equal to the averaged uncertainty in the data. Parameter estimates from the Superior Province (Darbyshire et. al., 2007) show the lowest correlation in $R$ as well as the highest mean difference in both parameters. These differences are well over the averaged computed error, and can not be easily explained by deviation related to noise in the data.

@@

We now proceed to compare our results computed using the ZK approach with those from the full grid search. We note that the velocity ratio $R$ has a correlation between the two datasets of 0.76 whereas $H$ has a correlation of 0.51. The low correlation between crustal thickness estimates can be explained by the strong tradeoff between $H$ and $V_P$ through the quantity

@@

\noindent As a consequence, contours of the misfit surface $s(H,R,V_P)$ are highly extended along the kinematic curve in the $H$ - $V_P$ space (Figure \ref{fig:kcurves} a). This trade-off contributes to the low correlation in $H$ between the two methods. Higher correlation in this quantity (0.96) can be restored by dividing out $V_P$ from $H$. The trade-off between the other parameters is lower as there is greater dependence between them such that a small change in one necessitates a large change in an other to maintain travel-time equality, which has the result of a well defined misfit surface (Figure \ref{fig:kcurves} b-f).

@@

There is less correlation between the crustal thickness data, though this is largely attributed to an approximate 10 km difference between the ARVN estimates. The reason for this discrepancy is unknown though both the ZK method and and a published study (Thompson et. al., 2010) provide an estimate of near 41 km for ARVN, much nearer the full grid value of 42 km. Despite lower correlation the apparent bias towards higher values computed from teleseismic data is also maintained for $H$. A positive bias in one parameter will tend to bias the other, owing to the relationship between thickness, velocity and times of a scattered phase. For a constant $t_{ps}$ one may compute how much $H$ must change for a 0.25 km/s bias in $V_P$ (Zhu and Kanamori, 2000). Using an average of the parameter estimates from the two cleanest stations, ULM and WHY, $\delta H$ will be 1.3 km, this fits well with the actual bias in the values.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{composition.eps}
  \caption{Comparison of major crustal elements from different studies. All estimates have been normalized to the results computed in this study.}
  \label{fig:composition}
\end{figure}

>>>>

\documentclass[review]{elsarticle}
\usepackage{graphicx,multirow,color,natbib,amssymb}
\usepackage[margin=1in]{geometry}

\makeatletter
\renewcommand\@biblabel[1]{}
\renewenvironment{thebibliography}[1]
     {\section*{\refname}%
      \@mkboth{\MakeUppercase\refname}{\MakeUppercase\refname}%
      \list{}%
           {\leftmargin0pt
            \@openbib@code
            \usecounter{enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\makeatother

\journal{Tectonophysics}

@@

\title{Seismic velocity estimation and the Canadian crust}

@@

\address[ubc]{Earth and Ocean Sciences, UBC, Vancouver BC, Canada}
\address[gsc]{Geological Survey Canada, Natural Resources Canada, Ottawa ON, Canada}
%% ------------------------------------------------------------------------ %%

\begin{abstract}
It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurements are compared with results for matching stations from previous studies. Analysis of results in conjunction with additional velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.
\end{abstract}

@@

The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and southern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6 Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98 Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean to Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occurred as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale d\'{e}collement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

Our knowledge of the continental crust in Canada, in particular its thickness and structure in depth, has advanced greatly through Lithoprobe, a now complete national geoscience project spearheaded by active-source seismic profiling (Clowes et al., 1999). The full suite of Lithoprobe seismic profiles form a near-continuous transect from the Pacific to the Atlantic Oceans, and has enabled many insights into crustal structure and genesis. Recent summary studies utilizing Lithoprobe data include the analysis of {\it Eaton} (2005) which considers seismic observations in light of several hypotheses for the origin of the continental Moho: relic Moho, underplating, metamorphic front and regional d\'{e}collement. Seismic evidence of preserved island arc Moho in Cascadia supports the first hypothesis whereas the presence of high-velocity, lower-crustal xenoliths in southern Alberta favours thick mafic underplating. Eclogite mineral assemblages forming a deep crustal root below the Grenville Front lends support to the third hypothesis, that at least in some places, continental Moho forms by large-scale eclogitization of the lower crust. Finally, seismic imaging of subhorizontal reflections separated by dipping reflection fabrics beneath the southern Canadian Cordillera suggests decoupling of the upper crust from the mantle during deformation and is taken as evidence for the fourth hypothesis, that an overprinted Moho may serve as a subhorizontal regional d\'ecollement. A subsequent study by {\it Cook et al.} (2009) argues that the continental Moho has varied and complex origins and that its imprint evolves through the effects of temperature, pressure, deformation and fluid migration.

In another recent summary study of Lithoprobe data, Hammer and Clowes (2011) discuss broad structural characteristics of the Canadian crust and note that Archean cratons and their boundaries contain structures similar to those found in Proterozoic and Phanerozoic orogens. They also discuss the lack of evidence for significant crustal disruption from magmatic intrusion related to subduction, post-orogenic extension or orogenic collapse. The authors remark upon the striking uniformity in Moho depth that is imaged between 33 km and 43 km depth along 20,000 km of Lithoprobe transects, despite the diversity of overlying crust. Both {\it Cook et al.} (2009) and {\it Hammer and Clowes} (2011) raise the possibility that, in some circumstances, the Moho and the base of seismic reflectivity do not represent the petrological crust-mantle boundary but rather eclogitized lower crust that is seismically indistinguishable from the upper mantle. If this is the case, the seismic Moho would represent the top of a metamorphic front and would be shallower than the crust-mantle boundary.

The Lith5.0 revised crustal model presented by {\it Perry et al.} (2002) was developed from crustal thickness estimates taken from Lithoprobe seismic profiles. This work updates the global Crust5.1 model (Mooney et al, 1998) and introduces broad crustal features of the Canadian crust such as relatively thin Archean Shield regions (32.5-40.0 km) as compared to the thick ($>$42.5 km) Phanerozoic crust in the south and west of the continental craton.

In this study, we examine the additional constraints that analyses of teleseismic receiver functions place upon Canadian bulk crustal properties when combined with available active source results. It is widely appreciated that receiver functions are sensitive primarily to crustal thickness and Poisson's ratio (or equivalently $V_P/V_S$) (Zhu and Kanomori, 2000), and that the latter quantity provides some ability to discriminate among possible bulk compositions (Zandt and Ammon, 1995). The constraint on Poisson's ratio and the greater averaging performed by low frequency (0.04-3 Hz) scattered teleseismic waves versus high frequency (10-40 Hz) {\it P}-waves employed in active source studies render the two techniques somewhat complementary. Our data set comprises the majority of teleseismic data available for the Canadian landmass including both permanent and temporary deployments. Many of the temporary deployments are described by {\it Bostock et al.} (2010). Three previous studies employing similar methodologies but with more regional emphases deserve special mention: an examination of crustal thickness and $V_P/V_S$ variation in the Grenville Orogen (Eaton et al., 2005), a structural and compositional characterization of the Superior Province (Darbyshire et al. 2006) and a crustal evolution analysis of the Hudson Bay region (Thompson et al., 2010). Results from these studies will serve as benchmarks for our expanded effort. In the following sections we describe the data and methods employed to interrogate the Canadian crust. We proceed to present the major results beginning with a comparison between our seismic parameter estimates and the estimates from three studies mentioned above for identical seismic stations. We then present the results of a modified approach to our parameter estimation method and follow this with a presentation of the resulting crustal property estimates.  First presented are the average crustal discontinuity and structural patterns across Canada. Second, we take a tour through those geological provinces for which sufficient data exists, focusing on regional variation and characterization. We conclude with a discussion on the implications for our understanding of the composition and evolution of the Canadian crust.

@@

Our analysis of bulk Canadian continental crust is based on estimates of crustal properties that can be accessed via seismic techniques, namely seismic wave velocities (or velocity ratio) and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source is teleseismic receiver function (hereafter RF) analysis. The remaining two datasets are Crust 1.0, a widely published statistical average of similar regions with a global one degree resolution, and a compilation of preprocessed active source data (Mooney, 2012). Both Crust 1.0 and the active source compilation are used to qualify and compare with the primary dataset.

\subsection{Teleseismic Data Set}
The primary data utilized in this study are computed from teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30$^\circ$-100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 events are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

@@

\begin{figure}
  \centering
    \includegraphics[width=0.7\textwidth]{reflectedPhases.pdf}
  \caption{Schematic diagram illustrating geometry of phases for the velocity contrast representing the Moho}
  \label{fig:reflectedPhases}
\end{figure}

A well tested and widely published method for extracting the ratio of {\it P} velocity $V_P$ to {\it S} velocity $V_S$, hereafter referred to as $R$, and crustal thickness (or depth to Moho) $H$, is outlined by {\it Zhu and Kanamori} (2000), hereafter ZK. This method exploits the dependence of these parameters to the differential arrival times between the {\it S} wave reflected phases $Ps$, $PpPs$, $PsPs$, $PpSs$ and $PsSs$ and the direct {\it P} wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ are kinematic analogues, such that the energy for these two phases arrive simultaneously for a 1-D model. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

@@

The travel time equations, as employed by ZK, assume a value for crustal {\it P} wave velocity, $V_P$, that in practice will trade-off to some degree with crustal thickness $H$. In our implementation of the ZK approach, each station is assigned a $V_P$ value corresponding to the Crust 1.0 value for the corresponding $1^\circ$ containing cell. In the specific cases where data are being compared to previously published results for quality control, $V_P$  values are chosen to match the values in the published study.

@@

A limitation of the ZK approach is the requirement of an initial $V_P$ estimate. In principal we may compute the stacking function Eq. \ref{eq:stack}, as a function of all three seismic parameters, $s(H,R,V_P)$. Taking advantage of modern multicore systems with reasonable quantities of data (fifty to a few hundred RFs) and a moderate search space ($n^3$ parameter candidates where $n \approx 150$) the computational time makes this method easily scalable to hundreds of stations. This ``full grid'' approach is analyzed to determine the viability of recovering $V_P$.

@@

The Crust 1.0 model (Laske et al., 2013) employs many of these same active source data to provide regular, complete coverage across all of Canada. The model includes estimates of $V_P$ and $V_S$ information with depth at $1^\circ$ resolution in latitude and longitude. In addition to employing local active source data, Crust 1.0 is also constrained by statistically aggregated data from geographically distant but geologically similar regions. It therefore affords a somewhat independent dataset suited for comparison with the crustal estimates calculated in this study.

@@

Three published studies have provided parameter estimates for $R$ and $H$ for subsets of the seismic stations considered here. The most recent study (Thompson et al., 2010) focused on the northern Canadian Shield and employs both the linear stacking approach used by ZK as well as a phase weighted method. A 2007 study of the Superior Province (Darbyshire et al, 2007) and a study of the Grenville Orogen (Eaton et al., 2006) utilize the ZK + semblance approach outlined in section \ref{section:VpVsMethod}.  Comparisons between crustal estimates of $R$ for particular seismic stations are made for those values with corresponding standard errors of less than $\pm$0.06. This error threshold is chosen empirically as that which best partitions those receiver function stacks with visually recognizable energy along phase moveout curves from those where lack of data or poor quality make these moveout curves difficult to distinguish. For each study under comparison, data are reprocessed using the $V_P$ chosen by the study authors. For the 2006 and 2007 studies the ZK + semblance method is used whereas for the Thompson et al. study the linear stacking ZK approach is employed. Both correlation coefficient and mean difference are provided in (Table \ref{table:comparison}) as measures of similarity.

\begin{table}
  \begin{tabular}{ p{3.8cm} p{3cm} p{2cm} p{1.2cm} p{1.2cm} p{1.2cm} p{1.2cm}}
    & & & \multicolumn{2}{ c }{Corr. Coeff.} & \multicolumn{2}{ c }{RMS difference} \\
    \hline
    Study Authors & Region & Num. Stns &  $H$ & $R$  &$H$ (km)& $R$ \\
    \hline
    Thompson et al. (2010) & N. Canadian Shield & 29 & 0.97 & 0.70 & 0.78 & 0.017 \\
    Darbyshire et al. (2007) & Superior Province & 10 & 0.95 & 0.43 & 1.71 & 0.037 \\
    Eaton et al. (2006) & Grenville Orogen & 26 & 0.90 & 0.61 & 1.03 & 0.033 \\
    \hline
  \end{tabular}
  \caption{Comparison of $R$ and $H$ estimates with three published studies}
\label{table:comparison}

@@

All three studies reported $H$ estimates that correlate strongly with the crustal thickness parameters computed here. The data from the northern Canadian Shield shows strong correlation in both $H$ and $R$. The lower correlation observed in estimates of $R$ than $H$ is not unexpected as the range of $R$ is relatively narrow and small changes ($\approx 2\%$) are physically meaningful. A corollary of this is greater sensitivity of $R$ to differences in procedure such as the choice of deconvolution algorithm and selection and scaling of data. The results from {\it Thompson et al.} (2010) also correspond most closely to the parameter estimates computed here, with the mean difference for both parameters being less than half of the averaged bootstrap error computed in this study. Results published by {\it Eaton et al.} (2006) show reasonable correlations in both $R$ and $H$ and a mean difference for both parameters that is roughly equal to the averaged uncertainty in the data. Parameter estimates from the Superior Province (Darbyshire et al., 2007) show the lowest correlation in $R$ as well as the highest mean difference in both parameters. These differences are well over the averaged computed error, and can not be easily explained by deviation related to noise in the data.

@@

We now proceed to compare our results computed using the ZK approach with those from the full grid search. We note that the velocity ratio $R$ has a correlation between the two datasets of 0.76 whereas $H$ has a correlation of 0.51. The low correlation between crustal thickness estimates can be explained by the strong trade-off between $H$ and $V_P$ through the quantity

@@

\noindent As a consequence, contours of the misfit surface $s(H,R,V_P)$ are highly extended along the kinematic curve in the $H$ - $V_P$ space (Figure \ref{fig:kcurves} a). This trade-off contributes to the low correlation in $H$ between the two methods. Higher correlation in this quantity (0.96) can be restored by dividing out $V_P$ from $H$. The trade-off between the other parameters is lower as there is greater dependence between them such that a small change in one necessitates a large change in the other to maintain travel-time equality, which has the result of a well defined misfit surface (Figure \ref{fig:kcurves} b-f).

@@

There is less correlation between the crustal thickness data, though this is largely attributed to an approximate 10 km difference between the ARVN estimates. The reason for this discrepancy may be an anomalously low quality active source reading as both the ZK method and a published study (Thompson et al., 2010) provide an estimate of near 41 km for ARVN, much nearer the full grid value of 42 km. Despite lower correlation the apparent bias towards higher values computed from teleseismic data is also maintained for $H$. A positive bias in one parameter will tend to bias the other, owing to the relationship between thickness, velocity and times of a scattered phase. For a constant $t_{ps}$ one may compute how much $H$ must change for a 0.25 km/s bias in $V_P$ (Zhu and Kanamori, 2000). Using an average of the parameter estimates from the two cleanest stations, ULM and WHY, $\delta H$ will be 1.3 km, this fits well with the actual bias in the values.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{composition.eps}
  \caption{Comparisons of crustal composition from selected studies. All estimates have been normalized to the results computed in this study.}
  \label{fig:composition}
\end{figure}
